Synthesizing Symmetric Lenses

### 2018-07-11 papers/SymmetricLenses/main.tex
a pass through section 5

% begin synthesis
\section{Synthesis}
\label{sec:synthesis}
Algorithm~\ref{alg:synth-sym-lens} presents our synthesis algorithm at a high
level of abstraction. The input regular expressions are first converted into
stochastic regular expressions with \ToStochastic.  
%Once that conversion has
%taken place, the $\RXSearch$ algorithm executes. This algorithm itself is
%defined via five functions: (1) $\PCF{\PQ.Create}$, (2) $\PCF{\PQ.Pop}$, (3)
%$\PCF{Expand}$, (4) $\PCF{\PQ.Push}$, and (5) $\PCF{\Continue}$. First,
This pair of SREs is used to initialize a priority queue ($pq$).
Next, $\PCF{SynthSymLens}$
enters a loop that searches for likely lenses. The loop terminates when the
algorithm believes it is unlikely to find a better lens than the best one it has
found so far (a termination condition defined by $\PCF{\Continue}$). Within each
iteration of the loop, it:
\begin{itemize}
\item pops the next class ($S$, $T$) of lenses to
search off of the priority queue ($\PCF{\PQ.Pop}$),
\item executes $\GreedySynth$ to find a best lens in that class if
one exists ($\LC$),
using the examples $exs$ to filter out potential lenses that do not satisfy
the specification,
\item replaces $best$ with $\LC$, if $\LC$ has lower cost according
to our information theoretic metric, and
\item adds all of the SREs derived from rewriting $S$ and $T$ ($\Expand(S,T)$)
  to the priority queue with $\PCF{\PQ.Push}$. The priority of a SRE pair is the
  number of rewrites needed to derive the pair from the originals.
\end{itemize}
When the loop terminates, the search returns the globally best lens found
($best$).  Each element of this algorithm will be explained in
further depth in the following subsections.

\begin{algorithm}
  \caption{\SynthSymLens}
  \label{alg:synth-sym-lens}
  \begin{algorithmic}[1]
    \Function{SynthSymLens}{$\BRegex,\BRegexAlt,\Examples$}
    \State $\Regex \gets \Call{\ToStochastic}{\BRegex}$
    \State $\RegexAlt \gets \Call{\ToStochastic}{\BRegexAlt}$
    \State $\RXSearchState \gets \Call{\PQ.Create}{\Regex,\RegexAlt}$
    \State $\Best \gets \None$
    \While{$\Call{\Continue}{\RXSearchState,\Best}$}
    \State $(\Regex,\RegexAlt) \gets \Call{\PQ.Pop}{\RXSearchState}$
    \State $\LC \gets \Call{\GreedySynth}{\Examples,\Regex,\RegexAlt}$
    \If{$\Call{Cost}{\LC} < \Call{Cost}{\Best}$}
    \State $\Best \gets \LC$
    \EndIf
    \State $\Call{\PQ.Push}{\RXSearchState,\Expand(\Regex,\RegexAlt)}$
    \EndWhile
    \State \ReturnVal{best}
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Search Through Types}
The first synthesizer is one that searches through regular expression pairs.
This synthesizer comprises \Expand, \PQ, and \Continue. It proposes regular
expressions that are semantically, but not syntactically, equivalent to the
originals. In particular, this synthesizer performs the following rewrites:
\begin{center}
  \begin{tabular}{rcl}
    $\PRegexStar{\Regex}{\Probability}$
    & \Rewrite
    & \PRegexOr{\EmptyString{}}{(\RegexConcat{\Regex{}}{\PRegexStar{\Regex{}}{\Probability}})}{\Probability}\\

@@

Our priority queue proposes stochastic regular expressions generated from fewer
rewrites first. Only when there are no more proposed regular expressions derived
from $n$ rewrites will \PCF{\PQ.Pop} propose regular expressions derived from
$n+1$ rewrites.

@@

When $\Call{Continue}{pq}$ returns $\False$, \RXSearch stops proposing regular
expression pairs, and instead returns to the user the best lens found thus far.
Note this means that if \RXSearch finds a bijective lens, which has zero cost it
will immediately return.
%
\subsection{Stochastic DNF Regular Expressions}
Stochastic DNF regular expressions (\SDNFREabbrev) are a pseudo-normal form
for stochastic regular expressions.  Of all the star-semiring equations
required to relate two \SDNFREabbrevs, only those relating to Kleene star
(\EG, \emph{Unrollstar$_L$} and \emph{Unrollstar$_R$}) are needed to relate
two \SDNFREabbrevs, other axioms, like associativity, commutativity
or distributivity, are unneeded.  This important property makes it possible to
construct an efficient algorithm ($\GreedySynth$), directed
by the syntax of two \SDNFREabbrevs, to synthesize any lenses in the
same ``class'' (\IE, any lenses that have the same type when the
Kleene star axioms are unused).  This property, and the resulting
efficiency of $\GreedySynth$ is key to the effectiveness of our
core algorithms.

@@

Intuitively, stochastic DNF regular expressions are essentially stochastic
regular expressions with all concatenations fully distributed over all
disjunctions. As such, the language of a stochastic DNF regular expression is a
union of its subcomponents, the language of a stochastic sequence is the
concatenation of its subcomponents, and the language of a stochastic atom is the
iteration of its subcomponent. For
$\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)}$
to be a valid stochastic DNF regular expression, the probabilities must sum to
one ($\sum_{i=0}^n\Probability_i = 1$).

@@

As these DNF regular expressions are \emph{stochastic}, they are annotated with
probabilities to express a probability distribution, in addition to a language. 
\begin{center}
  \begin{tabular}{rcl}
    $\ProbabilityOf{\PRegexStar{\DNFRegex}{\Probability}}{\String}$
    & =
    & $\Sigma_n \Sigma_{\String = \String_1 \ldots \String_n}\Probability^n*(1-\Probability)*\Pi_{i=1}^n\ProbabilityOf{\DNFRegex}{\String_i}$\\

@@

We have a means of turning stochastic regular expressions into stochastic DNF
regular expressions. The conversion algorithm itself, written
$\ToDNFRegexOf{\Regex}$, is defined below.
\[
  \begin{array}{rcl}
    \ToDNFRegexOf{\String} & = & \DNFOf{(\SequenceOf{\String},1)}\\
    \ToDNFRegexOf{\emptyset} & = & \DNFOf{}\\
    \ToDNFRegexOf{(\PRegexStar{\Regex}{\Probability})} & = & \AtomToDNFOf{\PRegexStar{(\ToDNFRegexOf{\Regex})}{\Probability}}\\
    \ToDNFRegexOf{(\RegexConcat{\Regex_1}{\Regex_2})} & = & \ToDNFRegexOf{\Regex_1} \ConcatDNF \ToDNFRegexOf{\Regex_2}\\
    \ToDNFRegexOf{(\PRegexOr{\Regex_1}{\Regex_2}{\Probability})} & = & \ToDNFRegexOf{\Regex_1} \OrDNF_{\Probability} \ToDNFRegexOf{\Regex_2}\\
  \end{array}
\]
This conversion utilizes operators defined in
Figure~\ref{fig:dnf-regex-functions}. We have proved that this conversion
respects languages and probability distributions.

@@

\paragraph*{\ToStochastic} With stochastic DNF regular expressions and
\ToDNFRegex defined, it is easier to explain the function that converts regular
expressions into stochastic regular expressions, \ToStochastic. If users
specified stochastic regular expressions themselves (perhaps inferred from
example data), this step could be skipped,
though our implementation does not (yet) permit users to provide such
stochastic regular expressions. If $\Regex$ is a stochastic DNF regular
expression generated by \ToStochastic, then $\ToDNFRegexOf{\Regex} =
\DNFOf{(\Sequence_1,\frac{1}{n}) \DNFSep \ldots \DNFSep
  (\Sequence_n,\frac{1}{n})}$ for some sequences $\Sequence_1\ldots\Sequence_n$,
and every stochastic atom generated by \ToStochastic is
$\PRegexStar{\DNFRegex}{.8}$. In particular, $\ToDNFRegex$ generates regular
expressions whose DNF form gives equal probability to all sequence subcomponents
of the DNF regular expression., and gives iterations a .8 chance to continue
iterating. In our experience generating random strings from regular expressions,
these probabilities provide good distributions of strings---stars are iterated
4 times on average, and no individual choice in a series disjunctions is chosen
disproportionately often. As we have found these distributions generate
reasonable strings, we believe they describe the shape of data well.

@@

\subsection{\GreedySynth}
\label{subsec:greedy-synth}
The greedy algorithm is actually three algorithms, one that greedily finds
symmetric DNF lenses (\GreedySynth), one that greedily finds symmetric sequence
lenses (\GreedySeqSynth), and one that finds symmetric atom lenses
(\AtomSynth{}). These three algorithms are hierarchically structured:
\GreedySynth relies on \GreedySeqSynth, \GreedySeqSynth relies on \AtomSynth,
and \AtomSynth rlies on \GreedySynth. The structure of the algorithms mirrors
the structure of symmetric DNF lenses and \SDNFREabbrevs. We describe this
synthesis algorithm in the context of trying to find a lens between
\lstinline{"" | name.name*} and \lstinline{"" | name}.

\paragraph*{Symmetric DNF Lenses} Algorithm~\ref{alg:greedysynth} presents
\GreedySynth, a means to synthesize symmetric DNF lenses. The inputs to
\GreedySynth are a suite of input-output examples and a pair of stochastic DNF regular expressions.

First, \PCF{CannotMap} checks whether there is no possible lens satisfying the
examples. If so, \GreedySynth returns \None early. Otherwise, \GreedySynth finds
the best lenses (given the examples that match them) between all sequence pairs
$(SQ_i, TQ_j)$ drawn from the left and right DNF regular expressions. (The
function \CartesianMap maps its argument across the cross product of the input
lists). A priority queue containing these sequence lenses, ordered by cost, is
then initialized with $\PCF{PQ.Create}$. The symmetric lens is then built up
iteratively from these sequence lenses, where the state of the partially
constructed lens is tracked in the lens builder, $lb$.

\GreedySynth loops until there are no more sequence lenses in the priority
queue. Within this loop, a sequence lense is popped from the queue and, if it is
``useful'', then it is included in the final DNF lens. The lens is considered to
be useful when its source (or target) is not already the source (or target) of
an already chosen sequence lens. If the examples require that two sequences have a
lens between them, then any sequence lens between them is considered useful. We
update the priorities of the sequence lenses as the algorithm proceeds: if two
sequence lenses have the same source, the second one to be popped gets a higher
cost than it originally had, to account for the information that needs to be stored
for including that source of non-bijectivity.

@@

\begin{algorithm}
  \caption{\GreedySynth}
  \label{alg:greedysynth}
  \begin{algorithmic}[1]
    \Function{GreedySynth}{$\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}$}
    \If{$\Call{CannotMap}{\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}}$}
    \State $\ReturnVal{\None}$
    \EndIf
    \State $\SLS \gets
    \Call{\CartesianMap}{\GreedySeqSynth(\Examples),\ListOf{\Sequence_1;\ldots;\Sequence_n},\ListOf{\SequenceAlt_1;\ldots;\SequenceAlt_m}}$
    \State $\GreedyState \gets \Call{PQ.Create}{\SLS}$
    \State $lb \gets \Call{LensBuilder.Empty}{}$
    \While{$\Call{PQ.IsNonempty}{\GreedyState}$}
    \State $\SSQLens \gets
    \Call{PQ.Pop}{\GreedyState}$
    \If{$\Call{LensBuilder.UsefulAdd}{lb,\SSQLens,\Examples}$}
    \State $lb \gets \Call{LensBuilder.AddSeq}{lb,\SSQLens}$
    \EndIf
    \EndWhile
    \State \ReturnVal{\Call{LensBuilder.ToDNFLens}{\GreedyState}}
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\paragraph*{Symmetric Sequence lenses} Algorithm~\ref{alg:greedyseqsynth}
presents \GreedySeqSynth, which synthesizes symmetric sequence lenses using an
algorithm whose structure is similar to \GreedySynth{}'s.  It calls \PCF{AtomSynth}, a
function that synthesizes atom lenses by iterating a DNF lens between
its subcomponents.

@@

\GreedySeqSynth then loops until there are no more atom lenses in the priority
queue. Within this loop, atom lenses are popped from the queue and the
``useful'' ones are kept. An atom lens is useful if adding it will lower the
cost of the generated sequence lens, or if examples show that one of the atoms
must not be disconnected. Each atom can only be part of one lens at a time, so
the algorithm must at times disconnect previously connected atoms to connect an
atom that must not be disconnected.

Finally, after all atoms have been popped, the sequence lens is validated and
returned. The algorithm confirms that all atoms that must not be disconnected
are involved in an atom lens. If they are not, \PCF{SLensBuilder.ToDNFLens}
instead returns \None.

\begin{algorithm}
  \caption{\GreedySeqSynth}
  \label{alg:greedyseqsynth}
  \begin{algorithmic}[1]
    \Function{AtomSynth}{$\Examples,\PRegexStar{\DNFRegex}{\Probability},\PRegexStar{\DNFRegexAlt}{\ProbabilityAlt}$}
    \If{$\Call{CannotMap}{\Examples,\PRegexStar{\DNFRegex}{\Probability},\PRegexStar{\DNFRegexAlt}{\ProbabilityAlt}}$}
    \State $\ReturnVal{\None}$
    \Else
    \State $\ReturnVal{\IterateLensOf{\GreedySynth(\Examples,\DNFRegex,\DNFRegexAlt)}}$
    \EndIf
    \EndFunction

@@

Our implementation includes a number of optimizations to this algorithm. We
added ``closed'' annotations to regular expressions \saz{what do these mean? I
  guess ``open'' means ``subject to DNF conversion'' and ``closed'' means
  not?}\afm{I describe in greater detail below, should I add in additional info
  here too?}, an expansion inference algorithm, hash-consing and memoization,
along with compositional synthesis. The implementation also allows an additional
parameter in specifications that customizes the termination condition for the
synthesis procedure.

\paragraph*{Open and Closed Regular Expressions} While \GreedySynth acts
relatively efficiently, it suffers from an exponential blowup when converting
regular expressions to DNF form. This can be minimized by not converting
\emph{all} of the regular expression into DNF form, but only converting
\emph{some} of it to DNF form, and doing so incrementally. This change is easy
to make by adding non-DNF stochastic regular expressions as a new type of
``closed'' atom. When converted into a DNF regular expression, if $\Regex$ is
annotated as ``closed'', then $\ToDNFRegexOf{\Regex} = \AtomToDNFOf{\Regex}$.
Regular expressions that are not annotated as closed, are considered ``open.''

At the start of synthesis, all regular expressions are closed. \RXSearch is
modified so that it is now permitted to rewrite closed regular expressions to
open ones (thereby triggering DNF normalization). \GreedySynth is allowed to to
map closed regular expressions to each other, but only using the identity
lens---they must parse the same string components in all examples.

\paragraph*{Expansion Inference} The prior optimization helped \GreedySynth, but
it made \RXSearch have the harder job of determining when to open closed
expressions.  To alleviate this difficulty, we added in expansion inference,
which identifies when a regular expression cannot possibly map by any means
other than the disconnect lens.  Expansion inference opens those regular
expressions to allow for further exploration.

\paragraph*{Hash-consing and Memoization} We further help \GreedySynth by
memoizing its results.  However, our stochastic DNF regular expressions are
quite large and so hashing them is computationally intensive.  We use less
memory and allow for quick comparisons and hashing of these DNF regular
expressions by hash-consing them.

\paragraph*{Compositional Synthesis} Compositional synthesis allows us to use
previously defined lenses---whether they are previously defined by users or by
synthesis. We alter \GreedySynth such that two regular expression atoms can be
mapped to each other either if they are the same regular expression and agree
with the examples, or if there is a lens which is a composition of predefined
lenses between them, where the lens agrees with the examples.

\paragraph*{Termination Customization}
In this work, we are constrained by three competing constraints. (1) We want to
return lenses quickly to users, (2) we want to return good lenses, and (3) we
don't want to require users to write a large number of examples in the
specifications. Our termination heuristic works well for most situations, but
there are situations where a large number of expansions must be performed to
find the correct lens.

Users could prevent incorrect lenses with a large number of examples, but it can
be quite difficult to write a large number of examples for complex formats. We
find it simpler to let the users specify that \RXSearch should be more
restrictive in which lenses it accepts, so users do not have to write a lot of
examples.

This constraint goes the other way too. If the user's desired transformation has
a particularly high cost, the algorithm may spend a long time searching in vain
for lenses with lower cost. By tuning \RXSearch to me more permissive in the
lenses it accepts, this search can be short-circuited.

For these types of situations, users can input a custom number, $t$, to
customize the strictness of the termination condition. In particular, if
$\PCF{PQ.Peek}(pq) = (\Regex,\RegexAlt)$, that RE pair is at distance $d$, the
number of pairs in $pq$ at distance $d$ is $n$, and the current best lens has
cost $c$:
\begin{gather*}
  \Call{Continue}{pq} := c > \Min(d + \Log_2(n) - t, 0)
\end{gather*}
With a positive $t$, the termination condition is more strict, and the algorithm
will spend longer searching for more likely lenses. With a negative $t$ the
termination condition is less strict, and the algorithm will spend less time
searching for more likely lenses.  Our default $t$ is set to $0$.

While our defaults termination condition quickly finds the correct lens for
\afm{35} of our \afm{45} benchmarks, our termination condition is not
one-size-fits all. Particularly complex lenses, or lenses that project a
particularly large amount of data, can be synthesized more easily and quickly by
customizing how permissive \RXSearch is in returning lenses.
% end synthesis

>>>>

% begin synthesis
\section{Synthesis}
\label{sec:synthesis}
Algorithm~\ref{alg:synth-sym-lens} presents our synthesis algorithm at a high
level of abstraction. The input regular expressions are first converted into
stochastic regular expressions with \ToStochastic.  
%Once that conversion has
%taken place, the $\RXSearch$ algorithm executes. This algorithm itself is
%defined via five functions: (1) $\PCF{\PQ.Create}$, (2) $\PCF{\PQ.Pop}$, (3)
%$\PCF{Expand}$, (4) $\PCF{\PQ.Push}$, and (5) $\PCF{\Continue}$. First,
This pair of SREs is used to initialize a priority queue ($pq$).
Next, $\PCF{SynthSymLens}$
enters a loop that searches for likely lenses. The loop terminates when the
algorithm believes it is unlikely to find a better lens than the best one it has
found so far (a termination condition defined by $\PCF{\Continue}$). Within each
iteration of the loop, it:
\begin{itemize}
\item pops the next class ($S$, $T$) of lenses to
search off of the priority queue ($\PCF{\PQ.Pop}$),
\item executes $\GreedySynth$ to find a best lens in that class if
one exists ($\LC$),
using the examples $exs$ to filter out potential lenses that do not satisfy
the specification,
\item replaces $best$ with $\LC$, if $\LC$ has lower cost according
to our information theoretic metric, and
\item adds the SREs derived from rewriting $S$ and $T$ ($\Expand(S,T)$)
  to the priority queue.  The priority of a SRE pair is the
  number of rewrites needed to derive the pair from the originals.
\end{itemize}
When the loop terminates, the search returns the globally best lens found
($best$).  Each subroutine of this algorithm will be explained in
further depth in the following subsections.

\begin{algorithm}[t]
  \caption{\SynthSymLens}
  \label{alg:synth-sym-lens}
  \begin{algorithmic}[1]
    \Function{SynthSymLens}{$\BRegex,\BRegexAlt,\Examples$}
    \State $\Regex \gets \Call{\ToStochastic}{\BRegex}$
    \State $\RegexAlt \gets \Call{\ToStochastic}{\BRegexAlt}$
    \State $\RXSearchState \gets \Call{\PQ.Create}{\Regex,\RegexAlt}$
    \State $\Best \gets \None$
    \While{$\Call{\Continue}{\RXSearchState,\Best}$}
    \State $(\Regex,\RegexAlt) \gets \Call{\PQ.Pop}{\RXSearchState}$
    \State $\LC \gets \Call{\GreedySynth}{\Examples,\Regex,\RegexAlt}$
    \If{$\Call{Cost}{\LC} < \Call{Cost}{\Best}$}
    \State $\Best \gets \LC$
    \EndIf
    \State $\Call{\PQ.Push}{\RXSearchState,\Expand(\Regex,\RegexAlt)}$
    \EndWhile
    \State \ReturnVal{best}
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Search Through Types}
The first synthesizer is one that searches through regular expression pairs.
This synthesizer comprises \Expand, \PQ, and \Continue. It proposes regular
expressions that are semantically, but not syntactically, equivalent to the
originals by performing the following rewrites:
\begin{center}
  \begin{tabular}{rcl}
    $\PRegexStar{\Regex}{\Probability}$
    & \Rewrite
    & \PRegexOr{\EmptyString{}}{(\RegexConcat{\Regex{}}{\PRegexStar{\Regex{}}{\Probability}})}{\Probability}\\

@@

The priority queue yields stochastic regular expressions generated using fewer
rewrites first. Only when there are no more proposed regular expressions derived
from $n$ rewrites will \PCF{\PQ.Pop} propose regular expressions derived from
$n+1$ rewrites.

@@

When $\Call{Continue}{pq}$ returns $\False$, \Expand stops proposing regular
expression pairs, and instead returns to the user the best lens found thus far.
Note this means that if the algorithm finds a bijective lens, which has zero cost, it
will immediately return.
%
\subsection{Stochastic DNF Regular Expressions}
Stochastic DNF regular expressions (\SDNFREabbrev) are a pseudo-normal form
for stochastic regular expressions.  Of all the star-semiring equations
required to relate two \SDNFREabbrevs, only those relating to Kleene star
(\EG, \emph{Unrollstar$_L$} and \emph{Unrollstar$_R$}) are needed to relate
two \SDNFREabbrevs; other axioms, like associativity, commutativity
or distributivity, are unneeded.  This important property makes it possible to
construct an efficient algorithm ($\GreedySynth$), directed
by the syntax of two \SDNFREabbrevs, to synthesize any lenses in the
same ``class'' (\IE, any lenses that have the same type when the
Kleene star axioms are unused).  This property, and the resulting
efficiency of $\GreedySynth$ is key to the effectiveness of our
core algorithms.

@@

Intuitively, stochastic DNF regular expressions are stochastic
regular expressions with all concatenations fully distributed over all
disjunctions. As such, the language of a stochastic DNF regular expression is a
union of its subcomponents, the language of a stochastic sequence is the
concatenation of its subcomponents, and the language of a stochastic atom is the
iteration of its subcomponent. For
$\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)}$
to be a valid stochastic DNF regular expression, the probabilities must sum to
one ($\sum_{i=0}^n\Probability_i = 1$).

@@

As these DNF regular expressions are \emph{stochastic}, they are annotated with
probabilities to express a probability distribution, in addition to a
language. \saz{I find the use of asterisk for the $\times$ or $\cdot$ or maybe
  just juxtaposition slightly jarring.}
\begin{center}
  \begin{tabular}{rcl}
    $\ProbabilityOf{\PRegexStar{\DNFRegex}{\Probability}}{\String}$
    & =
    & $\Sigma_n \Sigma_{\String = \String_1 \ldots \String_n}\Probability^n*(1-\Probability)*\Pi_{i=1}^n\ProbabilityOf{\DNFRegex}{\String_i}$\\

@@

The algorithm for converting a stochastic regular expressions into its
corresponding stochastic DNF regular expressions, written
$\ToDNFRegexOf{\Regex}$, is defined below.
\[
  \begin{array}{rcl}
    \ToDNFRegexOf{\String} & = & \DNFOf{(\SequenceOf{\String},1)}\\
    \ToDNFRegexOf{\emptyset} & = & \DNFOf{}\\
    \ToDNFRegexOf{(\PRegexStar{\Regex}{\Probability})} & = & \AtomToDNFOf{\PRegexStar{(\ToDNFRegexOf{\Regex})}{\Probability}}\\
    \ToDNFRegexOf{(\RegexConcat{\Regex_1}{\Regex_2})} & = & \ToDNFRegexOf{\Regex_1} \ConcatDNF \ToDNFRegexOf{\Regex_2}\\
    \ToDNFRegexOf{(\PRegexOr{\Regex_1}{\Regex_2}{\Probability})} & = & \ToDNFRegexOf{\Regex_1} \OrDNF_{\Probability} \ToDNFRegexOf{\Regex_2}\\
  \end{array}
\]
This conversion relies on operators defined in
Figure~\ref{fig:dnf-regex-functions}. We have proved that this conversion
respects languages and probability distributions.

@@

\paragraph*{\ToStochastic} With stochastic DNF regular expressions and
\ToDNFRegex defined, it is easier to explain \ToStochastic, the function that converts regular
expressions into stochastic regular expressions. If users
provided stochastic regular expressions themselves (perhaps inferred from
example data), this step could be skipped,
though our implementation does not (yet) permit users to provide such
stochastic regular expressions. If $\Regex$ is a stochastic DNF regular
expression generated by \ToStochastic, then $\ToDNFRegexOf{\Regex} =
\DNFOf{(\Sequence_1,\frac{1}{n}) \DNFSep \ldots \DNFSep
  (\Sequence_n,\frac{1}{n})}$ for some sequences $\Sequence_1\ldots\Sequence_n$,
and every stochastic atom generated by \ToStochastic is
$\PRegexStar{\DNFRegex}{.8}$. In particular, $\ToDNFRegex$ generates regular
expressions whose DNF form gives equal probability to all sequence subcomponents
of the DNF regular expression., and gives iterations a .8 chance to continue
iterating. In our experience generating random strings from regular expressions,
these probabilities provide good distributions of strings---stars are iterated
4 times on average, and no individual choice in a series disjunctions is chosen
disproportionately often. As we have found these distributions generate
reasonable strings, we believe they describe the shape of data well.

@@

\subsection{\GreedySynth}
\label{subsec:greedy-synth}
The synthesis procedure comprises three algorithms: one that greedily finds
symmetric DNF lenses (\GreedySynth), one that greedily finds symmetric sequence
lenses (\GreedySeqSynth), and one that finds symmetric atom lenses
(\AtomSynth{}). These three algorithms are hierarchically structured:
\GreedySynth relies on \GreedySeqSynth, \GreedySeqSynth relies on \AtomSynth,
and \AtomSynth rlies on \GreedySynth. The structure of the algorithms mirrors
the structure of symmetric DNF lenses and \SDNFREabbrevs. We describe this
synthesis algorithm using the running example of searching for a lens between
\lstinline{"" | name.name*} and \lstinline{"" | name}.

\paragraph*{Symmetric DNF Lenses} Algorithm~\ref{alg:greedysynth} presents
\GreedySynth, which synthesizes symmetric DNF lenses. Its inputs 
are a suite of input-output examples and a pair of stochastic DNF regular expressions.
%
First, \PCF{CannotMap} determines whether there is no lens satisfying the
examples, and if so, \GreedySynth returns \None immediately. Otherwise, \GreedySynth finds
the best lenses (given the examples that match them) between all sequence pairs
$(SQ_i, TQ_j)$ drawn from the left and right DNF regular expressions. (The
function \CartesianMap maps its argument across the cross product of the input
lists). A priority queue containing these sequence lenses, ordered by cost, is
then initialized with $\PCF{PQ.Create}$. The symmetric lens is then built up
iteratively from these sequence lenses, where the state of the partially
constructed lens is tracked in the lens builder, $lb$.

\GreedySynth loops until there are no more sequence lenses in the priority
queue. Within this loop, a sequence lense is popped from the queue and, if it is
``useful,'' it is included in the final DNF lens. The lens is considered to
be useful when its source (or target) is \textit{not} already the source (or target) of
an already chosen sequence lens. If the examples require that two sequences have a
lens between them, then any sequence lens between them is considered useful. We
update the priorities of the sequence lenses as the algorithm proceeds: if two
sequence lenses have the same source, the second one to be popped gets a higher
cost than it originally had, to account for the information that needs to be stored
for including that source of non-bijectivity.

@@

\begin{algorithm}[t]
  \caption{\GreedySynth}
  \label{alg:greedysynth}
  \begin{algorithmic}[1]
    \Function{GreedySynth}{$\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}$}
    \If{$\Call{CannotMap}{\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}}$}
    \State $\ReturnVal{\None}$
    \EndIf
    \State $\SLS \gets
    \Call{\CartesianMap}{\GreedySeqSynth(\Examples),\ListOf{\Sequence_1;\ldots;\Sequence_n},\ListOf{\SequenceAlt_1;\ldots;\SequenceAlt_m}}$
    \State $\GreedyState \gets \Call{PQ.Create}{\SLS}$
    \State $lb \gets \Call{LensBuilder.Empty}{}$
    \While{$\Call{PQ.IsNonempty}{\GreedyState}$}
    \State $\SSQLens \gets
    \Call{PQ.Pop}{\GreedyState}$
    \If{$\Call{LensBuilder.UsefulAdd}{lb,\SSQLens,\Examples}$}
    \State $lb \gets \Call{LensBuilder.AddSeq}{lb,\SSQLens}$
    \EndIf
    \EndWhile
    \State \ReturnVal{\Call{LensBuilder.ToDNFLens}{\GreedyState}}
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\paragraph*{Symmetric Sequence lenses} Algorithm~\ref{alg:greedyseqsynth}
presents \GreedySeqSynth, which synthesizes symmetric sequence lenses using an
algorithm whose structure is similar to \GreedySynth{}'s.  It calls \PCF{AtomSynth}, 
which synthesizes atom lenses by iterating a DNF lens between
its subcomponents.

@@

\GreedySeqSynth then loops until there are no more atom lenses in the priority
queue. Within the loop, a popped atom lens is considered 
``useful'' if adding it to the sequence will lower the
cost of the generated sequence lens, or if examples show that one of its atoms
must not be disconnected. Each atom can be part of only one lens at a time, so
the algorithm must sometimes a previously chosen atom lens in order to connect an
atom that must not be disconnected.
After all atoms have been popped, the sequence lens is validated and
returned. The algorithm succeeds when all atoms that must not be disconnected
are involved in an atom lens, but \PCF{SLensBuilder.ToDNFLens}
returns \None otherwise.

\begin{algorithm}[t]
  \caption{\GreedySeqSynth}
  \label{alg:greedyseqsynth}
  \begin{algorithmic}[1]
    \Function{AtomSynth}{$\Examples,\PRegexStar{\DNFRegex}{\Probability},\PRegexStar{\DNFRegexAlt}{\ProbabilityAlt}$}
    \If{$\Call{CannotMap}{\Examples,\PRegexStar{\DNFRegex}{\Probability},\PRegexStar{\DNFRegexAlt}{\ProbabilityAlt}}$}
    \State $\ReturnVal{\None}$
    \Else
    \State $\ReturnVal{\IterateLensOf{\GreedySynth(\Examples,\DNFRegex,\DNFRegexAlt)}}$
    \EndIf
    \EndFunction

@@

Our implementation includes a number of optimizations not described above:
annotations that guide DNF conversion; an expansion inference
algorithm; hash-consing and memoization; and compositional synthesis. The
implementation also provides for an additional specification parameter that
customizes the termination condition for the synthesis procedure.

\paragraph*{Open and Closed Regular Expressions} While \GreedySynth acts
relatively efficiently, it suffers from an exponential blowup when converting
regular expressions to DNF form. This can be mitigated by not converting
\emph{all} of the regular expression to DNF form, but only converting
\emph{some} of it, and doing so incrementally. This change is easy
to support by adding non-DNF stochastic regular expressions as a new type of
``closed'' atom. When converted into a DNF regular expression, if $\Regex$ is
annotated as ``closed'', then $\ToDNFRegexOf{\Regex} = \AtomToDNFOf{\Regex}$.
Regular expressions that are not annotated as closed, are considered ``open.''

At the start of synthesis, all regular expressions are closed. \Expand is
modified so that it is now permitted to rewrite closed regular expressions to
open ones (thereby triggering DNF normalization). \GreedySynth is allowed to to
map closed regular expressions to each other, but only by using the identity
lens---they must parse the same string components in all examples.

\paragraph*{Expansion Inference} Distinguishing open and closed regular
expressions helps \GreedySynth, but doing so means that \Expand has the harder job of
determining when to open closed expressions.  To that end,
expansion inference identifies when a regular expression cannot
possibly map by any means other than the ``trivial'' disconnect lens. Expansion inference
opens those regular expressions to allow for further exploration.

\paragraph*{Hash-consing and Memoization} We further improve \GreedySynth by
memoizing its results.  However, our stochastic DNF regular expressions are
quite large and so hashing them is computationally intensive.  We use less
memory and allow for quick comparisons and hashing of these DNF regular
expressions by hash-consing them.

\paragraph*{Compositional Synthesis} Compositional synthesis allows us to use
previously defined lenses---whether they are previously defined by users or by
synthesis. We alter \GreedySynth such that two regular expression atoms can be
mapped to each other either if they are the same regular expression and agree
with the examples, or if there is an existing lens (which is a composition of predefined
lenses) that already agrees with the examples.

\paragraph*{Termination Customization}
In this work, there are three competing constraints. (1) We want to return
lenses quickly to users, (2) we want to return good lenses, and (3) we don't
want to require users to write a large number of examples in their
specifications. Our termination heuristic works well for most situations, but
there are cases where a large number of expansions must be performed to find the
correct lens.  Users could work around that problem by providing a large number
of examples, but it can be difficult to write a large number of informative
examples for complex formats. We find it simpler to instead let the users
specify that \SynthSymLens should be more restrictive in which lenses it
accepts.  Conversely, if the user's desired transformation has a particularly
high cost, the algorithm may spend a long time searching in vain for lenses with
lower cost. By tuning \SynthSymLens to be more permissive, such a search can be
short-circuited.

For these situations, users can provide a number, $t$, which affects
the strictness of the termination condition, adjusting the \Continue formula
as follows. When $\PCF{PQ.Peek}(pq) = (\Regex,\RegexAlt)$, that RE pair is at
distance $d$, the number of pairs in $pq$ at distance $d$ is $n$, and the
current best lens has cost $c$:
\begin{gather*}
  \Call{Continue}{pq} := c > \Min(d + \Log_2(n) - t, 0)
\end{gather*}
With a positive $t$, the termination condition is stricter, and the algorithm
will spend longer searching for more likely lenses. With a negative $t$ the
termination condition is less strict, and the algorithm will spend less time
searching for more likely lenses.  Our default $t$ is $0$.

While our defaults termination condition quickly finds the correct lens for
\afm{35} of our \afm{45} benchmarks, our termination condition is not
one-size-fits all. Complex lenses, or lenses that project a
large amount of data, can be synthesized more easily and quickly by
adjusting $t$.
% end synthesis
